Explicit Scale Simulation for analysis of RNA-sequencing with ALDEx2

### 2025-03-11 go3.tex
submitted version Dec 6

\begin{document}
\maketitle
\begin{abstract}
In high-throughput sequencing (HTS) studies, sample-to-sample variation
in sequencing depth is driven by technical factors, and not by variation
in the scale (e.g., total size, microbial load, or total mRNA
expression) of the underlying biological systems. Typically a
statistical normalization is used to remove unwanted technical variation
in the data or the parameters of the model to enable analyses that are
sensitive to scale; e.g., differential abundance and differential
expression analyses. We recently showed that all normalizations make
implicit assumptions about the unmeasured system scale and that errors
in these assumptions can dramatically increase false positive and false
negative rates. We demonstrated that these errors can be mitigated by
accounting for uncertainty about scale using a \emph{scale model}, which
we integrated into the ALDEx2 R package. This article provides new
insights into those methods, focusing on the application to
transcriptomic analysis. Here we provide transcriptomic case studies
demonstrating how scale models, rather than traditional normalizations,
can reduce false positive and false negative rates in practice while
enhancing the transparency and reproducibility of analyses. We show that
these scale models replace the need for dual cutoff approaches often
used to address the disconnect between practical and statistical
significance. We demonstrate the utility of that scale models built
based on known housekeeping genes in complex metatranscriptomic
datasets. Thus this work provides example and practical guidance on how
to incorporate scale into transcriptomic analysis.
\end{abstract}

@@

\doublespacing

High-throughput sequencing (HTS) is a ubiquitous tool used to explore
many biological phenomenon such as gene expression (single-cell
sequencing, RNA-sequencing, meta-transcriptomics), microbial community
composition (16S rRNA gene sequencing, shotgun metagenomics) and
differential enzyme activity (selex, CRISPR killing). HTS proceeds by
taking a sample from the environment, making a library, multiplexing
(merging) multiple libraries together, and then applying a sample of the
multiplexed library to a flow cell. Each of these steps is a
compositional sampling step as only a fixed-size subsample of nucleic
acid is carried over to subsequent steps. Thus, with each sampling step
the connection between the size of the sampled DNA pool and the scale
(e.g., size, microbial load, or total gene expression) of the measured
biological system is degraded or lost. In the end, the information
contained in the data relates only to relative abundances and has an
arbitrary scale imposed by the sequencing process(1--3). Increasingly,
researchers are turning to modified experimental protocols in an attempt
to recover the lost biological variation in scale; see for example (4,
5). However, these protocols only recover scale variation at the
intermediate step in the sample preparation protocol where the
intervention was made. In short, the disconnect between sample-to-sample
variation in sequencing depth and biological variation in scale remains
an open challenge.

The analysis of HTS data suffers from several known problems that can be
traced, in whole or in part, to misspecification of scale. The first
issue is poor control of the false discovery rate (FDR) (6--9),
exhibited as dataset-dependent FDR control and by the disconnect between
statistical and biological significance(10). In current practice, these
issues are addressed by a dual-filtering method, whereby both a low
p-value (or equivalently a low q-value following FDR correction (11))
and a large difference between groups is used to identify interesting
transcripts or genes for follow-up analysis(10). This double-filtering
approach is graphically exemplified by the volcano plot (12), but is
known to not appropriately control the FDR\\
(13, 14). The second issue is poor performance when analyzing data where
the mean change between groups is non-zero. Such asymmetric data can
arise when a gene set is expressed in one group but not the other, or
when one group contains different gene content from the other. This type
of data frequently arises in in-vitro selection experiments (SELEX),
transcriptome analysis, and microbiome analysis (15). The third issue is
that these problems become more pronounced as more samples are
collected; that is, more information results in a worsening of the
accuracy of the analysis (16).

The three problems were recently shown by Nixon and Silverman (3) to be
a result of a mismatch between the underlying size or scale of the
system and the assumptions of the normalizations used for the analysis
of HTS. Biological variation in scale often represents an important
unmeasured confounder in HTS analyses (17). For example, cells
transformed by the cMyc oncogene have about 3 times the amount of mRNA
and about twice the rRNA content than non-transformed cells (18), and
this dramatically skews transcriptome analysis (19). In addition,
wild-type and mutant strains of cell lines, yeast or bacteria have
different growth rates and RNA contents under different conditions,
which affect our ability to identify truly differentially abundant genes
(20--22). As another example, the total bacterial load of the vaginal
microbiome differs by 1-2 orders of magnitude in absolute abundance
between the healthy and bacterial vaginosis states (23), and the
composition between these states is dramatically different (24, 25).
Thus, a full description of any of these systems includes both relative
change (composition) and absolute abundance (scale). Current methods
access only the compositional information yet make implicit assumptions
about the scale (16).

Recently, Nixon et al. (3) showed that the challenge of non-biological
variation in sequencing depth be viewed as a problem of
partially-identified models. They showed that \emph{all} normalizations
make some assumption about scale but these implicit assumptions are
often inappropriate and difficult to interpret. Because of this
different normalizations provide different outputs when applied to the
same dataset (6, 8, 26--28). Intuitively, normalizations in widespread
use assume that either all samples have the same scale,
e.g.~proportions, rarefaction (29), RPKM (30, 31), etc; or that a subset
of features in one sample can be chosen as a reference to which the
others are scaled e.g.~the TMM (32), or LVHA (15) or the additive
log-ratio (33); or that different sub-parts of each sample maintain a
constant scale across samples e.g.~the RLE (34); or that the geometric
mean of the parts is appropriate e.g the CLR (35) and its derivatives.

The original naive ALDEx2 (36) model unwittingly made a strict
assumption about scale through the CLR normalization (3). While this
assumption could be useful in many cases it could never be exactly true,
and others have shown that it is not always the best option (37). Nixon
et al. (3) showed that better scale assumptions resulted in more
reproducible data analysis including better control of both false
positive and false negative results. We recently modified ALDEx2 to
explicitly model the scale over a range of reasonable normalization
parameters, and showed significant improvements in performance in
microbiome and in-vitro selection experiments (16). Here, we briefly
introduce review modifications and then show how scale uncertainty can
greatly improve modeling in transcriptome and meta-transcriptome
datasets to provide more robust and reproducible results.

@@

Formal and expanded descriptions of the concepts that follow are given
in (3, 16). To be concrete, we let \(\mathbf{Y}\) denote the
\emph{measured} \(D \times N\) matrix of sequence counts with elements
\(\mathbf{Y}_{dn}\) indicating the number of measured DNA molecules
mapping to feature \(d\) (e.g., a taxon, transcript or gene) in sample
\(n\). Likewise, we can denote \(\mathbf{W}\) as the \emph{true} amount
of class \(d\) in the biological system from which sample \(n\) was
obtained. We can think of \(\mathbf{W}\) as consisting of two parts, the
scale \(\mathbf{W}^{\perp}\) (e.g., totals) and the composition
\(\mathbf{W}^{\parallel}\) (i.e., proportions). That is,
\(\mathbf{W}^{\perp}\) is a \(N\)-vector with elements
\(\mathbf{W}^{\perp}_{n}=\sum_{d}\mathbf{W}_{dn}\) while
\(\mathbf{W}^{\parallel}\) is a \(D \times N\) matrix with elements
\(\mathbf{W}^{\parallel}_{dn}=\mathbf{W}_{dn}/\mathbf{W}^{\perp}_{n}\).
Note that with these definitions \(\mathbf{W}\) can be written as the
element-wise combination of scale and composition:
\(\mathbf{W}_{dn}=\mathbf{W}^{\parallel}_{dn}\mathbf{W}^{\perp}_{n}\),
or as the logarithm
\(\log \mathbf{W}_{dn}= \log \mathbf{W}^{\parallel}_{dn} + \log \mathbf{W}^{\perp}_{n}\).

Many of the normalizations in widespread use in tools such as DESeq2
(38), edgeR (32), metagenomeSeq (39) ALDEx2 (40) can be stated as ratios
of the form
\(\hat{{\mathbf{W}}}_{dn} \approx \mathbf{Y}_{dn}/f(\mathbf{Y})\), where
the denominator is determined by some function of the observation. We
use the \(\hat{{\mathbf{W}}}\) (\(\ \hat{}\ \)) notation to indicate
that the output is an estimate of the true value. The technical
variation in sequencing depth
(\(\mathbf{Y}^{\perp}_{n}=\sum_{d}\mathbf{Y}_{dn}\)) implies that
observed data \(\mathbf{Y}\) provides us with information about the
system composition \(\mathbf{W}^{\parallel}\) but little to no
information in the system scale \(\mathbf{W}^{\perp}\) (Lovell et
al.~2011).

@@

The ALDEx2 R package (36, 40) is a general purpose toolbox to model the
uncertainty of HTS data and to use that model to estimate LFC
significance. At a high-level, ALDEx2 has three connected components to
estimate the uncertainty inherent in HTS datasets. First, the tool
accounts for the uncertainty of the sequencing counts using Dirichlet
multinomial sampling to build a probabilistic model of the data; i.e.,
\(\mathbf{\hat{W}}^{\parallel} \approx \mathrm{Dir}(\mathbf{Y})\).
Secondly, ALDEx2 uses the centred log-ratio transformation to scale the
data (36). However, this step was modified recently to account for scale
uncertainty and misspecification (16) via a scale model, explained with
more details in (3, 16) and summarized in the next paragraph. Finally, a
standard null-hypothesis test and a non-parametric estimate of mean
standardized difference are used to report on the finite sample
variation. These sources of uncertainty and variation are combined via
reporting the expected values from a Monte-Carlo simulation framework.
For simplicity, we use the term `difference' to refer to the absolute
difference between groups, and `dispersion' to refer to the
within-condition difference or pooled variance as defined in (36). These
are calculated on a \(\log_2\) scale. For more details on ALDEx2 see (3,
16, 36, 40).

Scale models can be incorporated into ALDEx2, turning the ALDEx2 model
into a specialized type of statistical model which Nixon et al. (3) term
a \textit{Scale Simulation Random   Variable} (SSRV). To do this, Nixon
et al. (3) generalized the concept of normalizations by introducing the
concept of a \textit{scale model} to account for potential error in the
centred log-ratio normalization step. They did this by including a model
for \(\mathbf{\hat{W}}^{\perp}_{n}\). The CLR method used by ALDEx2
makes the assumption \(\mathbf{\hat{W}}^{\perp}_{n}=1/G_{n}\), where
\(G_n\) is the geometric mean of sample n, which while being a random
variable, is essentially constant across each Monte-Carlo replicate.
With this modificaiton, ALDEx2 can be generalized by considering
probability models for the scale \(\mathbf{\hat{W}}^{\perp}_{n}\) that
have mean \(1/G_{n}\). For example, the following scale model
generalizes the CLR:

@@

This formulation is quite flexible (3, 16). In the simple or `default'
configuration, \(\mu = 0\) and \(\gamma\) is a tunable parameter drawn
from a log-Normal distribution that adds scale uncertainty. The default
approach controls only the degree of uncertainty in the CLR assumption
for the \(x_{n}\) binary condition indicator (e.g., \(x_{n}=1\) denotes
case and \(x_{n}=0\) denotes control). In the advanced or `informed'
configuration, \(\mu\) takes different values for each group, and this
controls the location of the LFC assumption. An example of both the
default and informed approaches is given for a microbiome dataset in
(16), and here we show that these approaches also work well in
transcriptome and metatranscriptome datasets. These modifications are
instantiated in ALDEx2 which is the first software package designed for
SSRV-based inference.

@@

Gierliński et al. (41) generated a highly replicated yeast transcriptome
dataset to compare gene expression between a wild-type strain and a snf2
gene knockout, \(\Delta\)snf2. This dataset was used to test several
RNA-seq tools for their power to detect the set of differentially
abundant transcripts identified in the full dataset when the data was
subset (10). In this study each tool had its own `gold standard' set of
transcripts with different tools identifying between between 65\% to
\textgreater80\% of all transcripts as being significantly different.
Since the majority of transcripts were significantly different, the
authors suggested that it was more appropriate to apply a dual cutoff
composed of both a Benjamini-Hochberg (42) corrected p-value (q-value)
plus a difference cutoff to limit the number of identified transcripts
to a much smaller fraction of the total. Nixon et al, (16) showed that
adding even a small amount of scale uncertainty with ALDEx2 dramatically
reduced the number of significant transcripts identified, removing the
need for the dual cutoff approach in this dataset and others. Below we
include an intuitive explanation of how incorporating scale uncertainty
achieved this outcome using a setting of \(\gamma=0.5\), which is in
line with the recommendations of (16) and that gives results comparable
to those proposed in (10).

We start with the assumption that not all statistically significant
differences are biologically relevant (43), and that such a large number
of significantly different transcripts breaks the necessary assumption
for DA/DE expression that most parts be invariant (27). As noted,
transcriptomics commonly uses a dual cutoff approach that is graphically
exemplified by volcano plots (10, 12). Using either DESeq2 or ALDEx2, a
majority of transcripts are statistically significantly different
between groups with a q-value cutoff of \(\le 0.05\); i.e.~4636 (79\%,
DESeq2) or 4172 (71\%, ALDEx2) of the 5891 transcripts. These values are
in line with those observed by (10). Such large numbers of statistically
significant transcripts seems biologically unrealistic. That 118
transcripts are identified by ALDEx2 and not DESeq2, while DESeq2
identifies 582 transcripts that ALDEx2 does not, suggests that the
choice of normalization plays a role in which results are returned as
significant and that some, if not the majority, are driven by technical
differences in the analysis (9, 27).

@@

The Volcano plots in Figure 1 A and B show that adding scale uncertainty
increases the minimum q-value and increases the concordance between the
q-value and the difference between groups (compare panels A and B). The
effect plots (44) in Figure 1C shows that the majority of significant
transcripts (red, orange) have negligible differences between groups and
very low dispersion. We suggest that this low dispersion is driven by
the experimental design which is actually a technical wet lab
replication rather than a true biological replication design (41). Scale
uncertainty can be incorporated using the \texttt{gamma} parameter that
controls the amount of noise added to the CLR mean assumption when we
call either \texttt{aldex()}, or \texttt{aldex.clr()}. Figure 1 B,D
shows that setting \(\gamma=0.5\) results in 205 which is far fewer
significant transcripts than in the naive analysis and we observe that
the minimum dispersion increases from 0.12 (\(\gamma = 0\) ) to 0.67
(\(\gamma=0.5\)).

@@

It is common practice to use a dual-cutoff by choosing transcripts based
on a thresholds for both q-values and fold-changes (10, 12). Note that
there is considerable variation in recommended cutoff values(10). Here,
applying a dual-cutoff using a heuristic of at least a \(2^{1.4}\) fold
change reduces the number of significant outputs to 193 for DESeq2 and
to 186 for ALDEx2. This cutoff was chosen for convenience and is in-line
with the recommendations of (10). the fold change limits are shown by
the dashed grey lines in Figure 1. The \(2^{1.4}\) fold change cutoff
identifies a similar number of transcripts as does ALDEx2 using
\(\gamma = 0.5\) which identifies 205. Supplementary Figure 2 shows how
to use the \texttt{aldex.senAnalysis()} function to identify those
transcripts that are very sensitive to scale uncertainty. In this
supplementary figure we see that even adding a very small amount of
scale \(\gamma = 0.1\) reduces the number of significant transcripts by
more than half. This allows us to those low-dispertion transcripts that
were significant only because of an absence of scale uncertainty. In
practice, we suggest that a \texttt{gamma} parameter between 0.5 and 1
is realistic for most experimental designs (16).

The effect on dispersion with increasing amounts of scale uncertainty
are shown in Figure 2A, where we can see that the dispersion increases
as uncertainty is added. Note that the dispersion in the unscaled
analysis in Figure 2A reaches a minimum near the mid-point of the
distribution, and also does so when the analysis is conducted with
DESeq2 (Supplementary Figure 3). This shows more clearly that dispersion
of many transcripts is almost negligible in the absence of scale
uncertainty. This plot makes the counter-intuitive suggestion that the
variance in expression of the majority of genes with moderate expression
is more predictable than highly-expressed genes or of housekeeping genes
(45). This is at odds with the known biology of cells where single cell
counting of highly-expressed transcripts shows that they have little
intrinsic variation (20, 46).

@@

Dos Santos et al. (47) used a vaginal metatranscriptome dataset to
compare the gene expression in bacteria collected from healthy (H) and
bacterial vaginosis (BV) affected women. In this environment, both the
relative abundance of species between groups and the gene expression
level within a species is different (48). Additionally, prior research
suggests that the total number of bacteria is about 10 times more in the
BV than in the H condition (23). Thus, these are extremely challenging
datasets in which to determine differential abundance as there are both
compositional and scale changes between conditions. The usual method to
analyze vaginal metratranscriptome data is to do so on an
organism-by-organism basis (48--50) because the scale confounding of the
environment is less pronounced. One attempt at system-wide analysis
returned several housekeeping functions as differentially expressed
between groups (49); a result likely due to a disconnect between the
assumptions of the normalization used and the actual scale of the
environment (15).

In this example, we show how to specify and interpret an informed scale
model that can explicitly account for some of these modeling
difficulties (16) even in a difficult dataset. An informed scale model
can control for both the mean difference of scale between groups (e.g.,
directly incorporate information on the differences in total number of
bacteria between the BV and H conditions) as well as the uncertainty of
that difference. To specify a user-defined scale model, we can pass a
matrix of scale values instead of an estimate of just the scale
uncertainty to \texttt{aldex.clr()}. This matrix should have the same
number of rows as the of Monte-Carlo Dirichlet samples, and the same
number of columns as the number of samples. While this matrix can be
computed from scratch by the analyst, there is an
\texttt{aldex.makeScaleModel()} function that can be used to simplify
this step in most cases. This encodes the scale model as
\(\Lambda \sim N(log_2 \mu_n, \gamma^{2})\), where \(\mu_n\) represents
the scale value for each sample or group and gamma is the uncertainty as
before. The scale estimate can be a measured value (cell count, nucleic
acid input, etc) or an estimate. Nixon et al. (16) showed that only the
ratio of the means are important when providing values for \(\mu_n\);
i.e., the ratio between the \(\log_2 \mu_i\) and \(\log_2 \mu_j\)
values. See the supplement to Nixon et al. (16) for more information.

>>>>

\begin{document}
\maketitle
\begin{abstract}
In high-throughput sequencing (HTS) studies, sample-to-sample variation
in sequencing depth is driven by technical factors, and not by variation
in the scale (size) of the biological system. Typically a statistical
normalization removes unwanted technical variation in the data or the
parameters of the model to enable differential abundance analyses. We
recently showed that all normalizations make implicit assumptions about
the unmeasured system scale and that errors in these assumptions can
dramatically increase false positive and false negative rates. We
demonstrated that these errors can be mitigated by accounting for
uncertainty using a \emph{scale model}, which we integrated into the
ALDEx2 R package. This article provides new insights focusing on the
application to transcriptomic analysis. We provide transcriptomic case
studies demonstrating how scale models, rather than traditional
normalizations, can reduce false positive and false negative rates in
practice while enhancing the transparency and reproducibility of
analyses. These scale models replace the need for dual cutoff approaches
often used to address the disconnect between practical and statistical
significance. We demonstrate the utility of scale models built based on
known housekeeping genes in complex metatranscriptomic datasets. Thus
this work provides guidance on how to incorporate scale into
transcriptomic data sets.
\end{abstract}

@@

\doublespacing 
\singlespacing

High-throughput sequencing (HTS) is a ubiquitous tool used to explore
many biological phenomenon such as gene expression (single-cell
sequencing, RNA-sequencing, meta-transcriptomics), microbial community
composition (16S rRNA gene sequencing, shotgun metagenomics) and
differential enzyme activity (selex, CRISPR killing). HTS proceeds by
taking a sample from the environment, making a library, multiplexing
(merging) multiple libraries together, and then applying a sample of the
multiplexed library to a flow cell. Each of these steps is a
compositional sampling step as only a fixed-size subsample of nucleic
acid is carried over to subsequent steps. Thus, with each sampling step
the connection between the actual size of the sampled DNA pool and the
scale (e.g., size, microbial load, or total gene expression) of the
measured biological system is degraded or lost. In the end, the
information contained in the data relates only to relative abundances
and has an arbitrary scale imposed by the sequencing process (1--3).
Researchers can use modified experimental protocols (4--7) or machine
learning methods (8) to uncover the biological variation in scale.
However, wet-lab protocols only provide information on the size of the
data downstream of the step in the sample preparation protocol where the
intervention was made and introduce an additional source of variation
that must be accounted for (5). The computational methods developed for
microbiome analysis have low correlation with the actual scale but are
useful (8). In short, the disconnect between sample-to-sample variation
in sequencing depth and biological variation in scale remains an open
challenge.

The analysis of HTS data suffers from several known problems that can be
traced, in whole or in part, to misspecification of scale. The first
issue is poor control of the false discovery rate (FDR) (9--13),
exhibited as dataset-dependent FDR control and by the disconnect between
statistical and biological significance (14). In current practice, these
issues are addressed by a dual-filtering method, whereby both a low
p-value (or equivalently a low q-value following FDR correction (15))
and a large difference between groups is used to identify interesting
transcripts or genes for follow-up analysis (14, 16). This
double-filtering approach is graphically exemplified by the volcano plot
(16), but is known to not appropriately control the FDR (17, 18). The
second issue is poor performance when analyzing data where the mean
change between groups is non-zero (3). Such asymmetric data can arise
when a gene set is expressed in one group but not the other, or when one
group contains different gene content from the other. This type of data
frequently arises in in-vitro selection experiments (SELEX),
transcriptome analysis, and microbiome analysis (19). The third issue is
that the actual scale of the environment is often a major confounding
variable during analysis (3, 8). The final issue is that these problems
become more pronounced as more samples are collected; that is, more
information results in a worsening of the accuracy of the analysis (3,
13, 20).

The four problems were recently shown by Nixon et al. (3) to be a result
of a mismatch between the underlying size or scale of the system and the
assumptions of the normalizations used for the analysis of HTS.
Biological variation in scale often represents an important unmeasured
confounder in HTS analyses (21). For example, cells transformed by the
cMyc oncogene have about 3 times the amount of mRNA and about twice the
rRNA content than non-transformed cells (22), and this dramatically
skews transcriptome analysis unless spike-in approaches are used (5). In
addition, wild-type and mutant strains of cell lines, yeast or bacteria
have different growth rates and RNA contents under different conditions,
which affect our ability to identify truly differentially abundant genes
(23--25). As another example, the total bacterial load of the vaginal
microbiome differs by 1-2 orders of magnitude in absolute abundance
between the healthy and bacterial vaginosis states (26), and the
composition between these states is dramatically different (27, 28).
Thus, a full description of any of these systems includes both relative
change (composition) and absolute abundance (scale). Current methods
access only the compositional information yet make implicit assumptions
about the scale (20).

Recently, Nixon et al. (3) showed that the challenge of non-biological
variation in sequencing depth be viewed as a problem of
partially-identified models. They showed that \emph{all} normalizations
make some assumption about scale but these implicit assumptions are
often inappropriate and difficult to interpret. This causes different
normalizations to provide different outputs when applied to the same
dataset (9, 11, 29--31). Intuitively, normalizations in widespread use
assume that either all samples have the same scale, e.g.~proportions,
rarefaction (32), RPKM (33, 34), etc; or that a subset of features in
one sample can be chosen as a reference to which the others are scaled
e.g.~the TMM (35), or LVHA (19) or the additive log-ratio (36); or that
different sub-parts of each sample maintain a constant scale across
samples e.g.~the RLE (37); or that the geometric mean of the parts is
appropriate e.g the CLR (38) and its derivatives.

The original naive ALDEx2 (39) model unwittingly made a strict
assumption about scale through the CLR normalization (3). This
assumption was often close enough to the true value to be useful, but
was not always the a good estimate and could be outperformed by other
normalizations (40). Nixon et al. (3) showed that better scale
assumptions resulted in more reproducible data analysis including better
control of both false positive and false negative results. We recently
modified ALDEx2 to explicitly model the scale over a range of reasonable
normalization parameters, and showed significant improvements in
performance in microbiome and in-vitro selection experiments (20). Here,
we briefly review these modifications and show how scale uncertainty can
greatly improve modeling in transcriptome and meta-transcriptome
datasets to provide more robust and reproducible results.

@@

Formal and expanded descriptions of the concepts that follow are given
in (3, 20). To be concrete, we let \(\mathbf{Y}\) denote the
\emph{measured} \(D \times N\) matrix of sequence counts with elements
\(\mathbf{Y}_{dn}\) indicating the number of measured DNA molecules
mapping to feature \(d\) (e.g., a taxon, transcript or gene) in sample
\(n\). Likewise, we can denote \(\mathbf{W}\) as the \emph{true} amount
of class \(d\) in the biological system from which sample \(n\) was
obtained. We can think of \(\mathbf{W}\) as consisting of two parts, the
scale \(\mathbf{W}^{\perp}\) (e.g., totals) and the composition
\(\mathbf{W}^{\parallel}\) (i.e., proportions). That is,
\(\mathbf{W}^{\perp}\) is a \(N\)-vector with elements
\(\mathbf{W}^{\perp}_{n}=\sum_{d}\mathbf{W}_{dn}\) while
\(\mathbf{W}^{\parallel}\) is a \(D \times N\) matrix with elements
\(\mathbf{W}^{\parallel}_{dn}=\mathbf{W}_{dn}/\mathbf{W}^{\perp}_{n}\).
Note that with these definitions \(\mathbf{W}\) can be written as the
element-wise combination of scale and composition:
\(\mathbf{W}_{dn}=\mathbf{W}^{\parallel}_{dn}\mathbf{W}^{\perp}_{n}\),
or as the logarithm
\(\log \mathbf{W}_{dn}= \log \mathbf{W}^{\parallel}_{dn} + \log
\mathbf{W}^{\perp}_{n}\).

Many of the normalizations in widespread use in tools such as DESeq2
(41), edgeR (35), metagenomeSeq (42) ALDEx2 (43) can be stated as ratios
of the form \(\hat{{\mathbf{W}}}_{dn}
\approx \mathbf{Y}_{dn}/f(\mathbf{Y})\), where the denominator is
determined by some function of the observation. We use the
\(\hat{{\mathbf{W}}}\) (\(\ \hat{}\
\)) notation to indicate that the output is an estimate of the true
value. The technical variation in sequencing depth
(\(\mathbf{Y}^{\perp}_{n}=\sum_{d}\mathbf{Y}_{dn}\)) implies that
observed data \(\mathbf{Y}\) provides us with information about the
system composition \(\mathbf{W}^{\parallel}\) but little to no
information in the system scale \(\mathbf{W}^{\perp}\) (Lovell et
al.~2011).

@@

The ALDEx2 R package (39, 43) is a general purpose toolbox to model the
uncertainty of HTS data and to use that model to estimate the underlying
LFC (log-fold change) significance. At a high-level, ALDEx2 has three
connected components to estimate the uncertainty inherent in HTS
datasets. First, the tool accounts for the uncertainty of the sequencing
counts using Dirichlet multinomial sampling to build a probabilistic
model of the data; i.e.,
\(\mathbf{\hat{W}}^{\parallel} \approx \mathrm{Dir}(\mathbf{Y})\).
Secondly, ALDEx2 uses the centred log-ratio transformation to scale the
data (39). However, this step was modified recently to account for scale
uncertainty and misspecification (20) via a scale model, explained with
more details in (3, 20) and summarized in the next paragraph. Finally, a
standard null-hypothesis test and a non-parametric estimate of mean
standardized difference are used to report on the finite sample
variation. These sources of uncertainty and variation are combined via
reporting the expected values from a Monte-Carlo simulation framework.
For simplicity, we use the term `difference' to refer to the absolute
difference between groups, and `dispersion' to refer to the
within-condition difference or pooled variance as defined in (39). These
are calculated on a \(\log_2\) scale. For more details on ALDEx2 see (3,
20, 39, 43).

Scale models can be incorporated into ALDEx2, turning the ALDEx2 model
into a specialized type of statistical model which Nixon et al. (3) term
a \textit{Scale Simulation Random   Variable} (SSRV). To do this, Nixon
et al. (3) generalized the concept of normalizations by introducing the
concept of a \textit{scale model} to account for potential error in the
centred log-ratio normalization step. They did this by including a model
for \(\mathbf{\hat{W}}^{\perp}_{n}\). The CLR normalization used by
ALDEx2 makes the assumption \(\mathbf{\hat{W}}^{\perp}_{n}=1/G_{n}\),
where \(G_n\) is the geometric mean of sample n, which while being a
random variable, is essentially constant across each Monte-Carlo
replicate, but that differs between samples. With this modification,
ALDEx2 can be generalized by considering probability models for the
scale \(\mathbf{\hat{W}}^{\perp}_{n}\) that have mean \(1/G_{n}\). For
example, the following scale model generalizes the CLR:

@@

This formulation is quite flexible (3, 20). In the simple or `default'
configuration, \(\mu = 0\) and \(\gamma\) is a tunable parameter drawn
from a log-Normal distribution(3). Adding scale uncertainty with the
\(\gamma\) paramenter controls only the degree of uncertainty of the CLR
assumption for the \(x_{n}\) binary condition indicator (e.g.,
\(x_{n}=1\) denotes case and \(x_{n}=0\) denotes control). In the
advanced or `informed' configuration, \(\mu\) takes different values for
each group and controls the location of the LFC assumption; combining
\(\mu\) with a \(\gamma\) estimate allows for uncertainty in both the
location and the scale. An example of both the default and informed
approaches is given for a microbiome dataset in (20) showing increased
sensitivity and specificity. Here we show that these approaches also
work well in transcriptome and metatranscriptome datasets. These
modifications are instantiated in ALDEx2 which is the first software
package designed for SSRV-based inference.

@@

Gierliński et al. (44) generated a highly replicated yeast transcriptome
dataset to compare gene expression between a wild-type strain and a snf2
gene knockout, \(\Delta\)snf2. This dataset was used to test several
RNA-seq tools for their power to detect the set of differentially
abundant transcripts identified in the full dataset when the data was
subset (14). In this study each tool had its own `gold standard' set of
transcripts with different tools identifying between between 65\% to
\textgreater80\% of all transcripts as being significantly different.
Since the majority of transcripts were significantly different, the
authors suggested that it was more appropriate to apply a dual cutoff
composed of both a Benjamini-Hochberg (45) corrected p-value (q-value)
plus a difference cutoff to limit the number of identified transcripts
to a much smaller fraction of the total. Nixon et al, (20) showed that
adding even a small amount of scale uncertainty with ALDEx2 dramatically
reduced the number of significant transcripts identified, removing the
need for the dual cutoff approach in this dataset and others. Below we
include an intuitive explanation of why and how incorporating scale
uncertainty achieved this outcome using a setting of \(\gamma=0.5\). The
approach is in line with the recommendations of (20) and gives results
comparable to those proposed in (14).

We start with the assumption that not all statistically significant
differences are biologically relevant (46), and that such a large number
of significantly different transcripts breaks the necessary assumption
for DA/DE expression that most parts be invariant (30). As noted,
transcriptomics commonly uses a dual cutoff approach that is graphically
exemplified by volcano plots (14, 16). Using either DESeq2 or ALDEx2, a
majority of transcripts are statistically significantly different
between groups with a q-value cutoff of \(\le 0.05\); i.e.~4636 (79\%,
DESeq2) or 4172 (71\%, ALDEx2) of the 5891 transcripts. These values are
in line with those observed by (14). Such large numbers of statistically
significant transcripts seems biologically unrealistic. That 118
transcripts are identified by ALDEx2 and not DESeq2, while DESeq2
identifies 582 transcripts that ALDEx2 does not, suggests that the
choice of normalization plays a role in which results are returned as
significant and that some, if not the majority, are driven by technical
differences in the analysis (13, 30).

@@

The Volcano plots in Figure 1 A and B show that adding scale uncertainty
increases the minimum q-value and increases the concordance between the
q-value and the difference between groups (compare panels A and B). The
effect plots (47) in Figure 1C shows that the majority of significant
transcripts (red, orange) have negligible differences between groups and
very low dispersion. We suggest that this low dispersion is driven by
the experimental design which is actually a technical wet lab
replication rather than a true biological replication design (44). Scale
uncertainty can be incorporated using the \texttt{gamma} parameter that
controls the amount of noise added to the CLR mean assumption when we
call either \texttt{aldex()}, or \texttt{aldex.clr()}. Figure 1 B,D
shows that setting \(\gamma=0.5\) results in 205 which is far fewer
significant transcripts than in the naive analysis and we observe that
the minimum dispersion increases from 0.12 (\(\gamma = 0\) ) to 0.67
(\(\gamma=0.5\)).

@@

It is common practice to use a dual-cutoff by choosing transcripts based
on a thresholds for both q-values and fold-changes (14, 16). Note that
there is considerable variation in recommended cutoff values(14). Here,
applying a dual-cutoff using a heuristic of at least a \(2^{1.4}\) fold
change reduces the number of significant outputs to 193 for DESeq2 and
to 186 for ALDEx2. This cutoff was chosen for convenience and is in-line
with the recommendations of (14) with the fold change limits shown by
the dashed grey lines in Figure 1. The \(2^{1.4}\) fold change cutoff
identifies a similar number of transcripts as does ALDEx2 using
\(\gamma = 0.5\) which identifies 205. Supplementary Figures 1 and 2
shows how to use the \texttt{aldex.senAnalysis()} function to identify
those transcripts that are very sensitive to scale uncertainty. In this
supplementary figure we see that even adding a very small amount of
scale \(\gamma = 0.1\) reduces the number of significant transcripts by
more than half. This allows us to ignore those low-dispersion
transcripts that were significant only because of an absence of scale
uncertainty. In practice, we suggest that a \texttt{gamma} parameter
between 0.5 and 1 is realistic for most experimental designs (20).

The effect on dispersion with increasing amounts of scale uncertainty
are shown in Figure 2A, where we can see that the dispersion increases
as uncertainty is added. Note that the dispersion in the unscaled
analysis in Figure 2A reaches a minimum near the mid-point of the
distribution, and also does so when the analysis is conducted with
DESeq2 (Supplementary Figure 3). This shows more clearly that dispersion
of many transcripts is almost negligible in the absence of scale
uncertainty. This plot makes the counter-intuitive suggestion that the
variance in expression of the majority of genes with moderate expression
is more predictable than highly-expressed genes or of housekeeping genes
(48). This is at odds with the known biology of cells where single cell
counting of highly-expressed transcripts shows that they have little
intrinsic variation (23, 49).

@@

Dos Santos et al. (50) used a vaginal metatranscriptome dataset to
compare the gene expression in bacteria collected from healthy (H) and
bacterial vaginosis (BV) affected women. In this environment, both the
relative abundance of species between groups and the gene expression
level within a species is different (51). Additionally, prior research
suggests that the total number of bacteria is about 10 times more in the
BV than in the H condition (26). Thus, these are extremely challenging
datasets in which to determine differential abundance as there are both
compositional and scale changes between conditions. The usual method to
analyze vaginal metratranscriptome data is to do so on an
organism-by-organism basis (51--53) because the scale confounding of the
environment is less pronounced. One attempt at system-wide analysis
returned several housekeeping functions as differentially expressed
between groups (52); a result likely due to a disconnect between the
assumptions of the normalization used and the actual scale of the
environment (19).

In this example, we show how to specify and interpret an informed scale
model that can explicitly account for some of these modeling
difficulties (20) even in a difficult dataset. An informed scale model
can control for both the mean difference of scale between groups (e.g.,
directly incorporate information on the differences in total number of
bacteria between the BV and H conditions) as well as the uncertainty of
that difference. To specify a user-defined scale model, we can pass a
matrix of scale values instead of an estimate of just the scale
uncertainty to \texttt{aldex.clr()}. This matrix should have the same
number of rows as the of Monte-Carlo Dirichlet samples, and the same
number of columns as the number of samples. While this matrix can be
computed from scratch by the analyst, there is an
\texttt{aldex.makeScaleModel()} function that can be used to simplify
this step in most cases. This encodes the scale model as
\(\Lambda \sim N(log_2 \mu_n,
\gamma^{2})\), where \(\mu_n\) represents the scale value for each
sample or group and gamma is the uncertainty as before. The scale
estimate can be a measured value (cell count, nucleic acid input, etc)
or an estimate. Nixon et al. (3, 20) showed that only the ratio of the
means are important when providing values for \(\mu_n\); i.e., the ratio
between the \(\log_2 \mu_i\) and \(\log_2 \mu_j\) values. See the
supplement to Nixon et al. (20) for more information.
